Julia for HPC: In Situ Data Analysis with Julia for Climate Simulations at Large Scale

### 2024-05-07 paper/paper.tex
Update paper.tex

Computing devices are evolving much faster than storage devices because of their different ratios of performance improvement to Research and Development (R\&D) investment in the post-Moore’s Law era~\cite{thompson2018decline}. This trend enlarges the performance gap between our computing and storage devices and greatly limits our ability to analyze all the generated scientific simulation data. When run on High-Performance Computing (HPC) systems, modern physics simulations employ thousands of nodes, generating large amounts of data. For example, at a high resolution, the three-dimensional physics model state alone can be tens of gigabytes \emph{per} time slice. While this data could possibly be compressed and then stored with high-performance input/output (I/O) for post-processing~\cite{jin2022accelerating,scorpio}, data movement and post-processing time can be extremely long. To address these issues, we adopt in situ data analysis to analyze physics simulation data as the simulation is running. The in situ data analysis design involves several design and performance tarde-offs in terms of coupling the physics simulation and data analysis, and we will explore some of the trade-offs in this paper.

Our target HPC application is Energy Exascale Earth System Model (E3SM)~\cite{golaz2019doe}. E3SM is the Department of Energy (DOE) state-of-the-science earth system model, and we focus on E3SM's atmosphere component (EAM) in this study. Developing data science techniques is extremely costly by using HPC programming languages such as C/C++ and Fortran due to the ecosystem lacking, so most data analysis programs are implemented by using Python. However, Python's performance is usually slow on HPC systems and we decided to use Julia as the programming language of developing data analysis modules because of its superior performance for data analysis and rich support of data analysis packages~\cite{hunold2020benchmarking}. Therefore, the major challenge of employing in situ data analysis to legacy scientific applications is that modern advanced data science techniques are usually not implemented in the same way as the HPC scientific codes. In this paper, we hypothesize that the Fortran-based EAM can be effectively coupled with in situ Julia-based data analysis modules on HPC systems.

To validate our hypothesis and demonstrate the in situ Julia data analysis infrastructure, we implement two in situ data analysis methods in Julia: (1) detection of Sudden Stratospheric Warming (SSW) with generalized extreme value (GEV) analysis, and (2) characterization of large-scale climate patterns via principal components analysis (PCA). SSW events are extreme events in the upper atmosphere (approximately 20km above ground level) that can cause the temperature to rise as much as 50$^o$C in only a few days.  This extreme warming can destabilize the polar vortex, which can in turn lead to rapid swings in surface air temperature. To identify SSWs, we developed an in situ analysis method that documents every occurrence when the zonal mean of the zonal wind becomes reversed (easterly) at 60°N and 10 hPa and lasts for at least 10 consecutive days in northern hemisphere winter. To characterize surface temperature changes that may result from SSWs, we fitted in situ generalized extreme value (GEV) models to the daily minimum surface temperatures at each spatial point in the continental United States (CONUS) via streaming variational inference~\cite{broderick2013streaming}. By updating two separate sets of models (one following SSWs and one not), we examined differences in extreme temperature behavior. Our second data analysis method focuses on PCA of the surface temperatures across the globe. PCA is a common statistical technique for identifying patterns in data that best explain variation in the data. Our TributaryPCA method estimates the top spatial principal component patterns in a streaming, distributed fashion~\cite{wang2021tributarypca}, whereas traditional analysis would need to save data from all spatial locations across many time steps to compute the principal components.

This paper presents a practical case study that uses Julia-based analysis in situ with a large-scale HPC application. The major benefit of this work is that it provides a reference design for coupling Julia with modern large-scale HPC applications which could be applied to other HPC applications with minimal effort.
In addition, data scientists could develop new Julia data analysis methods and insert them into the interface without recompiling the HPC code.
Another benefit of this work is an accurate and quantitative view of how the Julia infrastructure and coupling impact the original HPC application, which helps provide actionable guidance to HPC users before coupling Julia with their target HPC applications. To help HPC users adopt the in situ approach and Julia to meet their needs, this work provides the following key contributions:
\begin{enumerate}
\item We develop a novel in situ data analysis infrastructure for coupling E3SM with Julia for in situ data analysis at large-scale with high productivity.
\item We develop two in situ Julia data analysis methods for E3SM's EAM data at large-scale for identifying sudden stratospheric warmings and identifying dominant spatial patterns in temperature.
\item We evaluate the performance of this coupling infrastructure on a large-scale HPC system and study various design considerations, leading to guidelines for adopting Julia for in situ data analysis on other HPC applications.
\item We have open-sourced the code and instructions of reproducing this in situ data analysis with E3SM in \url{https://github.com/ltang85/In-Situ-data-Analysis-with-Julia-for-Climate-Simulations-at-Large-Scale/tree/main/src}.
\end{enumerate}

@@

Our work aims to provide data scientists with a performant interface for developing in situ data analysis algorithms without directly interacting with complex HPC codes or relying on post-hoc analysis. However, there are three challenges in doing so. First, E3SM's EAM is written in Fortran and adopts Message Passing Interface (MPI) for large-scale parallelism~\cite{walker1996mpi}, and there are no Fortran APIs for embedding Julia into EAM. Second, one MPI rank of EAM is only able to access to its local data (e.g., velocity and temperature), and each in situ Julia data analysis instance can only access the same local data of its parent EAM MPI rank. Last, EAM is compiled and interacting with other E3SM components, and the in situ Julia data analysis should be noninvasive to other E3SM components. To address these challenges, we develop and implement an infrastructure for coupling the Julia runtime with E3SM, which enables effective transfer of data and consistent MPI communication between Julia and E3SM through a minimalist design.

@@

Julia is a dynamically-typed programming language that is for general-purpose application development and provides many Python-like features (e.g., rich package support)~\cite{bezanson2017julia}. The first public Julia release was introduced in 2013 and it is designed to provide users with the speed of C/C++/Fortran~\cite{bezanson2017julia} while giving a convenient high-level programming environment. Julia is a built on top of the modularized LLVM compiler infrastructure~\cite{lattner2004llvm}, and features a Just-In-Time (JIT) compilation engine~\cite{aycock2003brief}. Therefore, Julia is ideal for agile development of advanced data science algorithms due to its speed and effectiveness in solving computationally complex problems. Also, Julia has support for parallelism and concurrency, which are crucial for HPC applications. In this paper, our major goal is to increase productivity by allowing data scientists to focus on writing high-level, performant numerical algorithms for legacy HPC applications at large-scale.

@@

Conventionally, large-scale HPC applications write their simulated scientific data to slow disks for post-hoc usage such as data analysis and visualization. As discussed before, the speed gap between computing and I/O is increasing, and modern supercomputers can generate large amounts of data that may not be fully stored. Therefore, the in situ approach is a processing technique that conducts the conventional post-hoc data analysis on the fly while the HPC application is still running, rather than using the data after it has been dumped to disk. That is, conventional disk I/O is converted to in-memory processing, which increases speed and enables analysis on full, rather than compressed and stored, simulation data. Employing this in situ technique enables novel data analysis methods that require large amounts of high-resolution simulation data from E3SM. Two conventional alternatives are data compression and high-performance I/O. For example, the Software for Caching Output and Reads for Parallel I/O (SCORPIO) library is used by E3SM for parallel data I/O~\cite{scorpio}.

@@

To couple E3SM's EAM with Julia, we implement a Fortran-based in situ interface in EAM to manage in situ Julia data analysis and pre-process the EAM simulation data. Specifically, this interface is implemented in EAM's \texttt{stepon} component, but we note this interface could be used for in situ data analysis on other E3SM components with minimal effort. Two major tasks are implemented in our Fortran interface. First, this interface calls the worker function of actual in situ Julia data analysis at every time step, and the initialization/cleanup functions at the first/last time step. Second, the interface wraps the EAM simulation data with its supportive data (e.g., data length, time step) and passes the pointer of the wrapped data to the C interface for removing explicit data copy. Also, this adapter passes an internal E3SM MPI communicator to the C interface. When the E3SM simulation is running at large-scale, each MPI rank of EAM calls its own Fortran interface and initiates one Julia instance. For the time steps between the first and last time step, each in situ Julia instance returns the control back to its parent EAM after it finishes its data analysis. After Julia returns its control back to EAM, the Julia runtime is still working and keeps global variables alive for usage in following time steps.

@@

The worker function in the C interface aims to support efficient low-level data communication between E3SM and the in situ Julia data analysis. The key implementation of this C interface is shown in the following code snippet. Each MPI rank of EAM has access to only a local data block of EAM variables (e.g., velocity and temperature) and passes its local data block's pointer (e.g., \texttt{arr} and \texttt{p\_arr} in the code snippet) to its own in situ Julia instance through the C interface. When the in situ Julia instance needs remote data (e.g., for computation of SSW) from other in situ Julia instances, \texttt{MPI.jl} is used in the Julia interface to implement the data communication between different in situ Julia instances. However, one key design challenge is to assure that E3SM and Julia use the same MPI communicator for correct data communication, as current Julia C embeddings are not able to directly pass the MPI communicator. To address this challenge, we transfer the original Fortran MPI communicator to this C interface without conversions, and then wrap it as \textit{jl\_box\_int32} and pass it to the Julia interface. The Fortran MPI communicator will be converted to Julia MPI communicator in the Julia interface. It is also important to root the Julia objects in this C interface between \textit{JL\_GC\_PUSH3()} and \textit{JL\_GC\_POP()}, preventing that the garbage collection (GC) frees the Julia references of \texttt{arr} and \texttt{p\_arr} in this C interface accidentally. This is because \texttt{arr} and \texttt{p\_arr} in the code snippet are allocated and deallocated in E3SM for every time step, and the Julia rooting can hold the references safely when running the Julia code.

@@

\subsection{Noninvasive Compilation}
The final design consideration is the compilation and execution design of coupling Julia with E3SM. The Fortran interface is implemented inside EAM, and the C interface file is placed in the same folder of the Fortran interface. The in situ Julia module is placed in the same run folder with the E3SM executable. As E3SM mixes the usage of GNU Make and CMake for combining and compiling different E3SM components, we have added the Julia compilation flags for the C and Fortran interfaces into the EAM CMake file (i.e., header files) and the top-level GNU Make file (i.e., Julia libraries). As we load the in situ Julia module via the Just-In-Time (JIT) scenario, the in situ Julia module is only compiled when it is called during runtime. Therefore, the whole infrastructure avoids recompiling the whole of E3SM when the in situ Julia module needs to be changed, and the in situ Julia module can be easily switched by only replacing the code of the same module file.

@@

Our first in situ climate analysis algorithm detects extreme weather events called SSWs. By definition, SSW is characterized as a major midwinter warming that occurs when the daily zonal mean zonal winds at $60^{\circ}$N and $10$ hPa (hectopascal) become easterly for at least $10$ consecutive days between November and March~\cite{ssw,ssw_def}. This event can result in extremely cold temperatures at Earth's surface, causing hazardous weather and disrupting many socioeconomic sectors. Since the robust detection of rare SSW events requires spatiotemporal climate data at high temporal frequencies, for high-resolution EAM runs, post-hoc SSW detection capabilities are limited as storage of high-resolution and high-frequency climate data would be prohibitive. Therefore, an in situ SSW detection algorithm is required.

Here, we briefly present the distributed SSW detection technique that we have developed and deployed in situ with EAM. During the simulation run, when a time marks the end of a day, we further process data from that time step. Initially, we compute the the zonal mean zonal wind at each MPI process independently using zonal velocity (U-velocity). As the EAM mesh does not put grid points exactly at $60^{\circ}$N and $10$ hPa, we first filter two layers of grid points, which are above and below the $10$ hPA pressure level and fall within [$59^{\circ}$N - $61^{\circ}$N] at each MPI process. Then, we linearly interpolate the zonal wind values to obtain values at $60^{\circ}$N and $10$ hPa. Then, using an MPI reduction, we estimate the final global daily zonal mean zonal wind value. Based on the definition of SSW, this value needs to be negative for at least $10$ consecutive days, so we maintain a global counter variable that keeps track of the number of consecutive days for which the value is negative. Once the value of that counter reaches $10$, we record a SSW event for that simulation year (where the event continues until the mean zonal wind becomes non-negative again). A pseudo code for this algorithm is provided in Algorithm~\ref{ssw_algo}. Note that when a SSW event is detected, the climate scientists typically are interested in collecting more data from following time steps, so the SSW detection can be used as a triggering event. Using SSW as a triggering event can reduce the overall in situ computation load since the SSW-specific analyses will only happen adaptively when a SSW event is detected.

\begin{algorithm}
\caption{In situ algorithm for SSW detection.}\label{ssw_algo}
\begin{algorithmic}[1]
\Require{EAM simulation data.} 
\Ensure{A list of time steps and corresponding simulation years when SSW is detected.}
\Statex
\For{each EAM time\_step}
  \If{(current\_time\_step == end\_of\_day)}
    \State Compute partial zonal mean zonal wind values
    \State at each MPI process.
    \State Compute the global zonal mean zonal wind
    \State values with MPI:Reduction.
    \If{(global zonal mean zonal wind < 0)}
    \State counter=counter+1
    \Else
    \State counter=0
    \EndIf
    \If{((counter == 10))}
    \State Record current time step as the beginning of
    \State a SSW event.
    %\State counter=0
    \EndIf
  \Else
    \State continue
  \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

To characterize spatial patterns in surface temperature across CONUS following a detected SSW (compared to no detected SSW), we fit two separate GEV models, similar to~\cite{ssw_isav}. These models represent the distribution of the daily minimum temperatures separately at each spatial location. Specifically, we transform the surface temperature data by subtracting 300$^\circ$K and multiplying by negative one, then fit GEV models. Briefly, we fit Gumbel models with probability density function $p(x; \mu, \beta) = \frac{1}{\beta}\exp \left\{ -\left[ \frac{x-\mu}{\beta}+\exp\left( -\frac{x-\mu}{\beta}\right)\right] \right\}$ where the initial priors are $p(\mu) \sim \mathcal{N}(0, 10)$ and $\beta \sim \text{LogNormal}(0, 0,5)$. As described in~\cite{ssw_isav}, Algorithm 1, model parameters are updated periodically using a buffer of recent data, depending on whether or not an SSW event was recently detected; at each step, the approximate variational posterior distribution from the previous step is used as the prior when updating the model with the new data point to achieve approximate streaming Bayesian inference across time points. In this paper, we update the GEV models every three days, using the last three days of daily minimum temperatures as the data for modeling. The fitted GEV models then represent distributional information about the data through their parameter estimates, which are updated in situ without the need to save the raw data. In this work, each time the GEV models were to be updated, we used 10 Monte Carlo samples in evaluating each gradient and 500 total optimization iterations each time the models were updated. We used the Adam optimizer from Turing~\cite{ge2018turing} with a learning rate of 0.0001.

>>>>

Computing devices are evolving much faster than storage devices because of their different ratios of performance improvement to Research and Development (R\&D) investment in the post-Moore’s Law era~\cite{thompson2018decline}. This trend enlarges the performance gap between our computing and storage devices and greatly limits our ability to analyze all the generated scientific simulation data. When run on High-Performance Computing (HPC) systems, modern physics simulations employ thousands of nodes, generating large amounts of data. For example, at a high resolution, the three-dimensional physics model state alone can be tens of gigabytes \emph{per} time slice. While this data could possibly be compressed and then stored with high-performance input/output (I/O) for post-processing~\cite{jin2022accelerating,scorpio}, data movement and post-processing time can be extremely long. To address these issues, we adopt in situ data analysis to analyze physics simulation data as the simulation is running. The in situ data analysis design involves several design and performance trade-offs in terms of coupling the physics simulation and data analysis, and we will explore some of the trade-offs in this paper.

Our target HPC application is Energy Exascale Earth System Model (E3SM)~\cite{golaz2019doe}. E3SM is the Department of Energy (DOE) state-of-the-science earth system model, and we focus on E3SM's atmosphere component (EAM) in this study. Developing data science techniques is extremely costly by using HPC programming languages such as C/C++ and Fortran due to the ecosystem lacking, so most data analysis programs are implemented by using Python. However, Python's performance is usually slow on HPC systems and we decided to use Julia as the programming language of developing data analysis modules because of its superior performance for data analysis and rich support of data analysis packages~\cite{hunold2020benchmarking}. Therefore, the major challenge of employing in situ data analysis in legacy scientific applications is that modern advanced data science techniques are usually not implemented in the same way as the HPC scientific codes. In this paper, we hypothesize that the Fortran-based EAM can be effectively coupled with in situ Julia-based data analysis modules on HPC systems.

To validate our hypothesis and demonstrate the in situ Julia data analysis infrastructure, we implement two in situ data analysis methods in Julia: (1) detection of Sudden Stratospheric Warming (SSW) with generalized extreme value (GEV) analysis, and (2) characterization of large-scale climate patterns via principal components analysis (PCA). SSW events are extreme events in the upper atmosphere (approximately 20km above ground level) that can cause the temperature to rise as much as 50$^o$C in only a few days.  This extreme warming can destabilize the polar vortex, which can in turn lead to rapid swings in surface air temperature. To identify SSWs, we developed an in situ analysis method that documents every occurrence when the zonal mean of the zonal wind becomes reversed (easterly) at 60°N and 10 hPa and lasts for at least 10 consecutive days in the northern hemisphere winter. To characterize surface temperature changes that may result from SSWs, we fitted in situ generalized extreme value (GEV) models to the daily minimum surface temperatures at each spatial point in the continental United States (CONUS) via streaming variational inference~\cite{broderick2013streaming}. By updating two separate sets of models (one following SSWs and one not), we examined differences in extreme temperature behavior. Our second data analysis method focuses on PCA of the surface temperatures across the globe. PCA is a common statistical technique for identifying patterns in data that best explain variation in the data. Our TributaryPCA method estimates the top spatial principal component patterns in a streaming, distributed fashion~\cite{wang2021tributarypca}, whereas traditional analysis would need to save data from all spatial locations across many time steps to compute the principal components.

This paper presents a practical case study that uses Julia-based analysis in situ with a large-scale HPC application. The major benefit of this work is that it provides a reference design for coupling Julia with modern large-scale HPC applications which could be applied to other HPC applications with minimal effort.
In addition, data scientists could develop new Julia data analysis methods and insert them into the interface without recompiling the HPC code.
Another benefit of this work is an accurate and quantitative view of how the Julia infrastructure and coupling impact the original HPC application, which helps provide actionable guidance to HPC users before coupling Julia with their target HPC applications. To help HPC users adopt the in situ approach and Julia to meet their needs, this work provides the following key contributions:
\begin{enumerate}
\item We develop a novel in situ data analysis infrastructure for coupling E3SM with Julia for in situ data analysis at large-scale with high productivity.
\item We develop two in situ Julia data analysis methods for E3SM's EAM data at large-scale for identifying sudden stratospheric warmings and identifying dominant spatial patterns in temperature.
\item We evaluate the performance of this coupling infrastructure on a large-scale HPC system and study various design considerations, leading to guidelines for adopting Julia for in situ data analysis on other HPC applications.
\item We have open-sourced the code and instructions for reproducing this in situ data analysis with E3SM in \url{https://github.com/ltang85/In-Situ-data-Analysis-with-Julia-for-Climate-Simulations-at-Large-Scale/tree/main/src}.
\end{enumerate}

@@

Our work aims to provide data scientists with a performant interface for developing in situ data analysis algorithms without directly interacting with complex HPC codes or relying on post-hoc analysis. However, there are three challenges in doing so. First, E3SM's EAM is written in Fortran and adopts Message Passing Interface (MPI) for large-scale parallelism~\cite{walker1996mpi}, and there are no Fortran APIs for embedding Julia into EAM. Second, one MPI rank of EAM is only able to access to its local data (e.g., velocity and temperature), and each in situ Julia data analysis instance can only access the same local data of its parent EAM MPI rank. Lastly, EAM is compiled and interacts with other E3SM components, and the in situ Julia data analysis should be noninvasive to other E3SM components. To address these challenges, we develop and implement an infrastructure for coupling the Julia runtime with E3SM, which enables effective transfer of data and consistent MPI communication between Julia and E3SM through a minimalist design.

@@

Julia is a dynamically typed programming language that is for general-purpose application development and provides many Python-like features (e.g., rich package support)~\cite{bezanson2017julia}. The first public Julia release was introduced in 2013 and it is designed to provide users with the speed of C/C++/Fortran~\cite{bezanson2017julia} while giving a convenient high-level programming environment. Julia is a built on top of the modularized LLVM compiler infrastructure~\cite{lattner2004llvm}, and features a Just-In-Time (JIT) compilation engine~\cite{aycock2003brief}. Therefore, Julia is ideal for agile development of advanced data science algorithms due to its speed and effectiveness in solving computationally complex problems. Also, Julia has support for parallelism and concurrency, which are crucial for HPC applications. In this paper, our major goal is to increase productivity by allowing data scientists to focus on writing high-level, performant numerical algorithms for legacy HPC applications at large-scale.

@@

Conventionally, large-scale HPC applications write their simulated scientific data to slow disks for post-hoc usage such as data analysis and visualization. As discussed before, the speed gap between computing and I/O is increasing, and modern supercomputers can generate large amounts of data that may not be fully stored. Therefore, the in situ approach is a processing technique that conducts the conventional post-hoc data analysis on the fly while the HPC application is still running, rather than using the data after it has been dumped to disk. That is, conventional disk I/O is converted to in-memory processing, which increases speed and enables dynamical analysis on full, rather than compressed and stored, simulation data. Employing this in situ technique enables novel data analysis methods that require large amounts of high-resolution simulation data from E3SM. Two conventional alternatives are data compression and high-performance I/O. For example, the Software for Caching Output and Reads for Parallel I/O (SCORPIO) library is used by E3SM for parallel data I/O~\cite{scorpio}.

@@

To couple E3SM's EAM with Julia, we implement a Fortran-based in situ interface in EAM to manage in situ Julia data analysis and pre-process the EAM simulation data. Specifically, this interface is implemented in EAM's \texttt{stepon} component, but we note this interface could be used for in situ data analysis on other E3SM components with minimal effort. Two major tasks are implemented in our Fortran interface. First, this interface calls the worker function of actual in situ Julia data analysis at every time step, and the initialization/cleanup functions at the first/last time step. Second, the interface wraps the EAM simulation data with its supportive data (e.g., data length, time step) and passes the pointer of the wrapped data to the C interface for removing explicit data copy. Also, this adapter passes an internal E3SM MPI communicator to the C interface. When the E3SM simulation is running at large-scale, each MPI rank of EAM calls its own Fortran interface and initiates one Julia instance. For the time steps between the first and last time step, each in situ Julia instance returns the control back to its parent EAM after it finishes its data analysis. After Julia returns its control back to EAM, the Julia runtime is still working and keeps global variables alive for usage in the following time steps.

@@

The worker function in the C interface aims to support efficient low-level data communication between E3SM and the in situ Julia data analysis. The key implementation of this C interface is shown in the following code snippet. Each MPI rank of EAM has access to only a local data block of EAM variables (e.g., velocity and temperature) and passes its local data block's pointer (e.g., \texttt{arr} and \texttt{p\_arr} in the code snippet) to its own in situ Julia instance through the C interface. When the in situ Julia instance needs remote data (e.g., for computation of SSW) from other in situ Julia instances, \texttt{MPI.jl} is used in the Julia interface to implement the data communication between different in situ Julia instances. However, one key design challenge is to ensure that E3SM and Julia use the same MPI communicator for correct data communication, as current Julia C embeddings are not able to directly pass the MPI communicator. To address this challenge, we transfer the original Fortran MPI communicator to this C interface without conversions, and then wrap it as \textit{jl\_box\_int32} and pass it to the Julia interface. The Fortran MPI communicator will be converted to Julia MPI communicator in the Julia interface. It is also important to root the Julia objects in this C interface between \textit{JL\_GC\_PUSH3()} and \textit{JL\_GC\_POP()}, preventing that the garbage collection (GC) frees the Julia references of \texttt{arr} and \texttt{p\_arr} in this C interface accidentally. This is because \texttt{arr} and \texttt{p\_arr} in the code snippet are allocated and deallocated in E3SM for every time step, and the Julia rooting can hold the references safely when running the Julia code.

@@

\subsection{Noninvasive Compilation}
The final design consideration is the compilation and execution design of coupling Julia with E3SM. The Fortran interface is implemented inside EAM, and the C interface file is placed in the same folder as the Fortran interface. The in situ Julia module is placed in the same run folder with the E3SM executable. As E3SM mixes the usage of GNU Make and CMake for combining and compiling different E3SM components, we have added the Julia compilation flags for the C and Fortran interfaces into the EAM CMake file (i.e., header files) and the top-level GNU Make file (i.e., Julia libraries). As we load the in situ Julia module via the Just-In-Time (JIT) scenario, the in situ Julia module is only compiled when it is called during runtime. Therefore, the whole infrastructure avoids recompiling the whole of E3SM when the in situ Julia module needs to be changed, and the in situ Julia module can be easily switched by only replacing the code of the same module file.

@@

Our first in situ climate analysis algorithm detects extreme weather events called SSWs. By definition, SSW is characterized as a major midwinter warming that occurs when the daily zonal mean zonal winds at $60^{\circ}$N and $10$ hPa (hectopascal) become easterly for at least $10$ consecutive days between November and March~\cite{ssw,ssw_def}. This event can result in extremely cold temperatures at the Earth's surface, causing hazardous weather and disrupting many socioeconomic sectors. Since the robust detection of rare SSW events requires spatiotemporal climate data at high temporal frequencies, for high-resolution EAM runs, post-hoc SSW detection capabilities are limited as storage of high-resolution and high-frequency climate data would be prohibitive. Therefore, an in situ SSW detection algorithm is required.

Here, we briefly present the distributed SSW detection technique that we have developed and deployed in situ with EAM. During the simulation run, when a time marks the end of a day, we further process data from that time step. Initially, we compute the zonal mean zonal wind at each MPI process independently using zonal velocity (U-velocity). As the EAM mesh does not put grid points exactly at $60^{\circ}$N and $10$ hPa, we first filter two layers of grid points, which are above and below the $10$ hPA pressure level and fall within [$59^{\circ}$N - $61^{\circ}$N] at each MPI process. Then, we linearly interpolate the zonal wind values to obtain values at $60^{\circ}$N and $10$ hPa. Then, using an MPI reduction, we estimate the final global daily zonal mean zonal wind value. Based on the definition of SSW, this value needs to be negative for at least $10$ consecutive days, so we maintain a global counter variable that keeps track of the number of consecutive days for which the value is negative. Once the value of that counter reaches $10$, we record an SSW event for that simulation year (where the event continues until the mean zonal wind becomes non-negative again). A pseudo code for this algorithm is provided in Algorithm~\ref{ssw_algo}. Note that when an SSW event is detected, the climate scientists typically are interested in collecting more data from the following time steps, so the SSW detection can be used as a triggering event. Using SSW as a triggering event can reduce the overall in situ computation load since the SSW-specific analyses will only happen adaptively when an SSW event is detected.

\begin{algorithm}
\caption{In situ algorithm for SSW detection.}\label{ssw_algo}
\begin{algorithmic}[1]
\Require{EAM simulation data.} 
\Ensure{A list of time steps and corresponding simulation years when SSW is detected.}
\Statex
\For{each EAM time\_step}
  \If{(current\_time\_step == end\_of\_day)}
    \State Compute partial zonal mean zonal wind values
    \State at each MPI process.
    \State Compute the global zonal mean zonal wind
    \State values with MPI:Reduction.
    \If{(global zonal mean zonal wind < 0)}
    \State counter=counter+1
    \Else
    \State counter=0
    \EndIf
    \If{((counter == 10))}
    \State Record current time step as the beginning of
    \State an SSW event.
    %\State counter=0
    \EndIf
  \Else
    \State continue
  \EndIf
\EndFor
\end{algorithmic}
\end{algorithm}

To characterize spatial patterns in surface temperature across CONUS following a detected SSW (compared to no detected SSW), we fit two separate GEV models, similar to~\cite{ssw_isav}. These models represent the distribution of the daily minimum temperatures separately at each spatial location. Specifically, we transform the surface temperature data by subtracting 300$^\circ$K and multiplying by negative one, then fit GEV models. Briefly, we fit Gumbel models with probability density function $p(x; \mu, \beta) = \frac{1}{\beta}\exp \left\{ -\left[ \frac{x-\mu}{\beta}+\exp\left( -\frac{x-\mu}{\beta}\right)\right] \right\}$ where the initial priors are $p(\mu) \sim \mathcal{N}(0, 10)$ and $\beta \sim \text{LogNormal}(0, 0,5)$. As described in~\cite{ssw_isav}, Algorithm 1, model parameters are updated periodically using a buffer of recent data, depending on whether or not an SSW event was recently detected; at each step, the approximate variational posterior distribution from the previous step is used as the prior when updating the model with the new data point to achieve approximate streaming Bayesian inference across time points. In this paper, we update the GEV models every three days, using the last three days of daily minimum temperatures as the data for modeling. The fitted GEV models then represent distributional information about the data through their parameter estimates, which are updated in situ without the need to save the raw data. In this work, each time the GEV models were to be updated, we used 10 Monte Carlo samples to evaluate each gradient and 500 total optimization iterations. We used the Adam optimizer from Turing~\cite{ge2018turing} with a learning rate of 0.0001.

\subsection{TributaryPCA} 
As described more fully in~\cite{wang2021tributarypca}, TributaryPCA solves the standard PCA objective in a distributed, online fashion; we give a brief overview of the implementation here.
Let $X \in \mathbb{R}^{d \times N}$ be a data set consisting of $N$ samples with unknown covariance $\Sigma \in \mathbb{R}^{d \times d}$. 
In our application, $N$ is the number of time points and $d$ is the number of spatial locations.
PCA seeks a $k$-dimensional orthonormal subspace $V \in \mathbb{R}^{d \times k}$ such that when the data are projected onto $V$, their variance is maximized. This amounts to solving 
\begin{equation}\label{eqn:pca-obj}
    \max_{\substack{V \in \mathbb{R}^{d \times k} \\ V^TV = I_k}} \text{Tr}\Big(V^T \Sigma V\Big).
\end{equation}
Equation~\eqref{eqn:pca-obj} is maximized by the top-$k$ eigenvectors of $\Sigma$, so the classical PCA calculation computes the top-$k$ eigenvectors of the estimated covariance matrix.
However, this approach is not feasible when data arrives sequentially (so that all $N$ samples are not available at one time) or when the dimension of the data is large (so that it is not stored centrally on one compute node).
In the streaming setting, the AdaOja algorithm~\cite{henriksen2019adaoja} seeks to estimate the solution to Equation~\eqref{eqn:pca-obj} using projected stochastic gradient descent with an automatically-tuned step size.
That is, the estimated eigenvectors $V$ are adjusted as each data point arrives (over time) using the gradient of Equation~\eqref{eqn:pca-obj} evaluated at the new data point.
This procedure consists of three key steps: a gradient computation, a learning rate update, and a QR decomposition (to ensure the orthogonality of the estimated eigenvectors).
The TributaryPCA algorithm (1) computes the gradient in a distributed fashion using data stored on each node and MPI \texttt{allreduce}, (2) updates the learning rate in a distributed fashion using MPI \texttt{allreduce}, (3) applies the stochastic gradient update on each node, and (4) uses distributed linear algebra~\cite{demmel2012communication} to orthogonalize the estimated $V$ (collected from all nodes using MPI \texttt{reduce}).
In streaming PCA, data $X_i \in \mathbb{R}^{d \times 1}, i=1,...,N$ arrives sequentially and the goal is to compute $V$ using all $N$ samples, where $V$ is updated sequentially with the current estimate denoted as $V_i$. 
The AdaOja algorithm relies on an initial learning rate $\alpha$ which we initialize to $1e-5$. In TributaryPCA, we have the additional complication that data $X_i$ is partitioned across $B$ compute nodes. 
The key portion of the in situ code (that would be called at each time step during the simulation) is shown in the following listing, where \texttt{X\_par} is the data on the current node (of shape $d_b \times 1$ where $d_B$ is the dimension of the data on a single compute node indexed by $b$), \texttt{V\_par} is the node's spatial subset of the current global estimate of the eigenvectors (of shape $d_B \times k)$, \texttt{comm} is the MPI communicator object, \texttt{master} denotes the root/master node, and \texttt{grad\_par}, \texttt{abs2}, $\alpha$ are pre-initialized arrays.
The code has three key steps: first, the gradient for the AdaOja streaming PCA update is computed; this quantity is $X_i X_i^T V_{i-1}$, which is computed with a local dot product at each node followed by an \texttt{Allreduce} to combine results across nodes, and another multiplication locally.
The second step is to update the learning rate based on the norm of the gradient and take a gradient step to update $V_{i-1}$ to $V_i$ based on data $X_i$.
Finally, the result must be projected to ensure $V_i$ is orthogonal.
This is accomplished by QR decomposition $V_i = Q_i R_i$ where $V_i$ is updated to the orthogonal $Q_i$. 
For distributed QR decomposition, we use CholeskyTSQR~\cite{demmel2012communication}.
